**Effective application of machine learning to microbiome based classification problems**

### 2019-10-14 submission/manuscript.Rmd
Jenna's edits 3rd round

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a reusable open-source pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and not immidiately interpretable. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable. Our analysis highlights the importance of choosing an ML approach should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

__Model selection and pipeline construction__ We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had a median AUROC value of `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`] which was higher than the other models for detecting SRNs. However, we determined that the random forest model performance was not statistically significantly higher (p = 0.5) when we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134]. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e., done using the test data).

@@

Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added [@westcott_opticlust_2017]. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. A grid search was needed to avoid large variation in prediction performance. Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016]. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization. For SVM with radial basis function kernel, we also tuned the sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has. For random forest, we tuned the number of features to consider when looking for the best tree split. For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners.

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models by 
taking the difference of AUROC values from the same datasplit. We plotted a histogram that showed the AUROC differences for 100 datasplits and calculated the empirical p-value of the histogram (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for double tail event (e.g., Figure S3).

@@

**Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network. We would like to thank the members of the Schloss lab for their valuable feedback during the formulation of the manuscript. Salary support for M.R came from NIH grant 1R01CA215574. Salary support for P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. The predictive performance of random forest model was higher than other ML models though not significantly (p=0.5). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that are associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that are associated with classifying a subject having an SRN had positive signs and were shown in red. Some of the same OTUs were identified as important in all of the linear models. 
\newpage
\includegraphics{Figure_4.png}

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.png}

@@

**Figure S1. Hyperparameter setting performances for linear models.** (A) L2 logistic regression, (B) L1 SVM with linear kernel, and (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model. 
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) Random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.  The stars represent the highest performing hyperparameter setting for the models. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

@@

**Figure S4. Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}

@@

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline

\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task? \\ Do we have a priori hypotheses? \\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem? \\ How much interpretability \\ does the problem need? } & \makecell[l]{Do we know the correlated \\ features? } \\ \hline

\makecell[l]{Classification \\ algorithm} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem? \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need? \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline

\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation?} & \makecell[l]{Do we have a held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline

\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters? } & \makecell[l]{Did we search the \\ full grid space and \\optimized our model? } \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs? } \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it? \\ Have we checked for \\ the effect of correlated \\ features?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results? } \\ \hline

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a reusable open-source pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and not immidiately interpretable. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable. Our analysis highlights the importance of choosing an ML approach based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

__Model selection and pipeline construction__. We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134]. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))` which indicated high degree of collinearity in the dataset. Permutation analysis allowed us to gauge the importance of each OTU in non-linear models and partially account for collinearity by grouping correlated OTUs to determine their impact as a group.

@@

Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking [@wiens_no_2019]. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added [@westcott_opticlust_2017]. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. A grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization. For SVM with radial basis function kernel, we also tuned the sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has. For random forest, we tuned the number of features to consider when looking for the best tree split. For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners. Performing a grid search for hyperparameter selection might not be feasible for when there are more than two hyperparameters to tune for. In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].)

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models pairwise by calculating the difference between AUROC values from the same datasplit (for 100 datasplits). We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).

@@

**Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network. We would like to thank the members of the Schloss lab for their valuable feedback. Salary support for M.R came from NIH grant 1R01CA215574. Salary support for P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. The predictive performance of random forest model was higher than other ML models,s though not significantly (p=0.5). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each datasplit. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. Some of the same OTUs were identified as important in all of the linear models. 
\newpage
\includegraphics{Figure_4.png}

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.png}

@@

**Figure S1. Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model. 
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.  The stars represent the highest performing hyperparameter setting for the models. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

@@

**Figure S4. Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}

@@

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline

\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application. \\ (e.g. input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline

\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline

\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g. when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline

\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g. feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
\end{tabular}
\newpage
